---
title: "Effects of Educational Television"
output: 
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \fancyhead[L]{}
  - \fancyhead[C]{This material should not be shared beyond those who are enrolled in this class}
---

In this exercise we're going to look at the effect 
of a educational television program 
[The Electric Company](https://en.wikipedia.org/wiki/The_Electric_Company) 
that ran from 1971-77 on children's reading scores.
We will investigate what reading gains, if any, were made by the 1st through 
4th grade classes as part of a randomized experiment.

This exercise is based on:

> Joan G. Cooney (1976) [The Electric Company: Television and Reading, 1971-1980: A Mid-Experiment Appraisal](https://files.eric.ed.gov/fulltext/ED130635.pdf). Children's Television Network Report.
The data comes from a two location trial in which 
treatment was randomized at the level of school classes.^[Classes were 
paired, but we will ignore that in the analysis] Each class was 
either treated (to watch the program) or control (to not watch the 
program). The outcome of interest is the score on a reading test 
administered at the end of each year called `post.score`. Note that 
these are distinct classes from all four years.  The variables are:

-------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `pair`               The index of the treated and control pair (ignored
                      here).
 
 `city`               The city: Fresno ("F") or Youngstown ("Y")
 
 `grade`              Grade (1 through 4)
 
 `supp`               Whether the program replaced ("R") or supplemented 
                      ("S") a reading activity

 `treatment`          "T" if the class was treated, "C" otherwise (randomized)
 
 `pre.score`          Class reading score *before* treatment, at the 
                      beginning of the school year
 
 `post.score`         Class reading score at the end of the school year
-------------------------------------------------------------------------------



## Question 1

Read the data into an data frame named `electric`. 
What sort of variable has R assumed `grade` is? How will 
it be treated in a linear model if we use it as an independent variable? 
Under what circumstances would that be reasonable or unreasonable?

Make a new variable from `grade` that is a factor. How will a 
linear model treat this new variable? 
**Hint:** You may 
find that `summary` illuminates the new data set.

Finally, overwrite the existing treatment variable so that it is 
numerical: 1 when the class is treated and 0 when not.

  
## Question 2

Let's now consider the effect of treatment. First, fit a linear model 
that predict `post.score` with just treatment. Then fit a model uses 
your factor version of `grade` as well as treatment.  

Summarise both models in terms of how much of the variance 
in `post.score` they "explain" and the median size of their 
errors. 

Now, consider each model's treatment coefficient. 
Are the estimates of this coefficient 
*different* in the two models? Why do you think that is?


## Question 3
Now make another model that uses the factor version of `grade` and 
`pre.score` (the reading score before the year begins) to predict 
`post.score`.  Is this model better? If so, in what ways?


## Question 4
Now let's consider the effect of treatment *within* each grade. 
We can use the `lm` function's `subset` argument to fit the model on just a 
subset of all the rows in the data set. For example, we can fit a model 
of the relationship of `post.score` to `treatment` and `pre.score` just 
in grade 2 like this:

```
mod <- lm(post.score ~ treatment + pre.score, data = electric, 
          subset = grade == 2)
```


Fit a linear model predicting `post.score` using treatment and 
`pre.score` for each 
grade. You need to follow the following procedures:

1. Define a function named `fit_reg` that returns the coefficient on treatment.
   The function should have two arguments: the entire data (`data_all`) and the grade (`grade_subset`).
2. Use a `for loop` and call the `fit_reg()` function for each grade (`1` to `4`).
   Store what the `fit_reg()` function returns in a variable.
3. Print out the coefficient on treatment using the `print()` function.
4. Briefly comment on the result. There are now *four* treatment effects. How do they differ as grade increases? 



## Question 5
Now let's try to learn about separate grade effects in a single model. One 
way to do this is to *interact* treatment with grade. Here's a general 
modeling principle: 

> If you think the *effect* of variable A varies according to the *values* of 
> variable B, then you should think of *adding an interaction* between A and B in 
> your model
Reminder: In the `lm` formula interface this amounts to adding an `A:B` term. For example,
if A and B interact to predict Y then the formula would be 
```{r, eval = FALSE}
Y ~ A + B + A:B
```
which would fit the model 
$$
Y_i = \beta_0 + \beta_A A_i + \beta_B B_i + \beta_{AB} (A_i \times B_i)  + \epsilon_i
$$
An alternative syntax to fit this model is `A*B`. So to fit the model above using this 
notation the formula is 
```{r, eval = FALSE}
Y ~ A * B 
```
Since we always want to have `A` and `B` if we have an `A:B` term, the `*` notation makes 
sure we don't forget any of them. But they are equivalent.

Fit a model of all the grades that includes `pre.score`, `treatment`, `grade` (factor version),
the factor version of grade interacted with `treatment`, and the factor version of grade interacted with `pre.score`
(this is called a fully interacted model). 
How would 
you construct grade-specific treatment effects from these coefficients?
Show an example for grade 2.



## Question 6
Use a bar plot to visualize the grade-specific treatment effects that
you calculated in the previous question. Briefly interpret the result.

\textbf{Hint:} You can make a bar plot using a `barplot()` function (textbook p.81)
```{r, warning=FALSE, message=FALSE, fig.align='center', fig.height = 3.5, fig.width = 3.5}
# Example
res <- data.frame(Y = c(1, 2, 3),
                  X = c("A", "B", "C"))
barplot(height = res$Y, names = res$X,
        xlab = "X", ylab = "Y",
        main = "Example")
```


